\section{ATM PERFORMANCE INDICATORS}

In the context of the performance-based management advocated by the International Civil Aviation Organization (ICAO), Key Performance Indicators (KPIs) are essential metrics used to quantitatively express past and current operational performance against organizational objectives. In the aviation industry, KPIs are crucial for enhancing governance and operational efficiency in a highly complex and regulated sector. By relying on standardized metrics, stakeholders can identify bottlenecks, measure outcomes, and promote continuous improvements in safety, punctuality, airport capacity, and environmental sustainability. 

Furthermore, the systematic analysis of KPIs facilitates communication among airport operators, airlines, air navigation service providers (ANSPs), and regulatory bodies, fostering a shared vision of strategic goals. KPIs also play a vital role in aligning aviation operations with international policies, such as the United Nations Sustainable Development Goals, by measuring environmental impacts like fuel consumption and emissions.

In this study, two specific performance indicators established by ICAO are of primary interest: KPI08 and KPI16. Both metrics are fundamental for evaluating the efficiency of Air Traffic Management (ATM) operations within the Terminal Maneuvering Area (TMA).

\subsection{KPI08: Additional Time in TMA}
KPI08 measures the additional transit time spent by an aircraft within the TMA compared to an optimal, unimpeded baseline. The reference baseline is generally established through the clustering of transit time samples based on the runway in use, the aircraft type, and the entry sector into the terminal area. The unimpeded transit time is then defined as the 20th percentile of all samples within these grouped conditions. 

Various factors can influence this indicator, including airspace design, meteorological conditions, and tactical interventions by Air Traffic Control (ATC) such as holding patterns and vectoring for sequencing. The comparison of the actual transit time to the 20th percentile baseline effectively isolates the operational inefficiencies, focusing on the ATM's responsibility in optimizing the arrival flow.

\subsection{KPI16: Additional Fuel Burn}
KPI16 represents the core metric for evaluating ATM-induced operational efficiency from an environmental perspective, providing an aggregate view of the excess fuel consumed during a flight phase. In the context of TMA operations, it is directly correlated with the inefficiencies captured by KPI08. 

The additional fuel burn is calculated as the difference between the actual fuel consumed by an aircraft and a reference fuel consumption. Similar to KPI08, this baseline is determined as the 20th percentile of fuel consumption for similar flights. Positive values of KPI16 indicate fuel consumption above the expected optimal level—often resulting from airspace congestion and ATC interventions—whereas negative values indicate flights that operated more efficiently than the established baseline. As the target variable of the predictive models developed in this work, KPI16 serves as a direct proxy for the environmental and economic cost of ATM inefficiencies in the terminal area.
