\section{RELATED WORKS}
\label{sec:related_works}

The aviation industry's growing commitment to sustainability and operational efficiency has spurred significant research into aircraft fuel consumption and emissions modeling. Recent advancements have increasingly leveraged machine learning and deep learning to overcome the limitations of traditional, static physics-based models. 

A substantial body of work focuses on micro-level or airframe-specific analyses. For instance, Jarry \textit{et al.} \cite{jarry2024generalization, jarry2024qar, jarry2025ageing} explored deep learning architectures to estimate fuel flow, emphasizing the generalization properties of these models across unseen aircraft types and specifically incorporating engine ageing effects using Quick Access Recorder (QAR) data. Similarly, deep learning frameworks utilizing Multi-Layer Perceptrons (MLP), Radial Basis Functions, and Recurrent Neural Networks (RNN) have been extensively proposed to predict fuel flow rates and model tail-specific aircraft fuel consumption, capturing individual airframe degradation trends \cite{baklacioglu2021fuel, ssrn2025tail}. While these studies excel in modeling the detailed physical performance and deterioration of individual engines, they rely heavily on high-fidelity, proprietary telemetry data (QAR). In contrast, our work abstracts from individual airframe degradation and focuses on macro-level operational inefficiencies caused by the Air Traffic Management (ATM) environment. By using aggregated radar data and standard base of aircraft data (BADA) performance tables, our model offers a highly scalable approach that Air Navigation Service Providers (ANSPs) can deploy without needing access to airline-specific telemetry.

At the operational and ATM level, Kwasiborska and Skorupski \cite{kwasiborska2021merging} assessed the environmental impact of merging landing aircraft streams, demonstrating how flow misalignment in the approach area increases holding times and subsequent fuel consumption. Krauth \textit{et al.} \cite{krauth2023generative} applied deep generative modeling (Variational Autoencoders) to simulate and generate 4-dimensional aircraft trajectories in Terminal Maneuvering Areas (TMAs) to estimate collision risks and account for ATC deviations. While Krauth \textit{et al.} focused on generating realistic trajectories, and Kwasiborska theoretically evaluated merging techniques, our study complements these perspectives by providing a concrete, data-driven regression framework to quantify the actual fuel penalties (KPI16) incurred during these TMA congestions and vectoring maneuvers. 

Finally, broader emission prediction models have also been developed. Al Kazzaz \cite{alkazzaz2025emissions} proposed an Artificial Neural Network (ANN) trained on the ICAO engine emissions databank to forecast CO2, NOx, and CO emissions based on engine and ambient parameters. While their approach aims to estimate direct chemical output for sustainability metrics, our methodology predicts the operational excess fuel burn in the TMA, which directly translates to excess emissions but is modeled from the perspective of airspace congestion and entry sector dynamics.

Overall, while existing literature heavily explores physics-based engine degradation and theoretical traffic flow models, our approach bridges the gap by using scalable, tree-based machine learning (specifically CatBoost) to quantify the real-world fuel cost of airspace inefficiencies, equipping stakeholders with a practical tool to optimize TMA operations dynamically.
